---
format: gfm
title: Measles models
---

## Highschool model

### Parameters

```{r}
yaml::read_yaml("highschool-params.yaml") |>
    str()
```

### Run

The following is the output from the highschool model:

```{r}
#| echo: false
system2("make", "run", wait = TRUE)
```

This model simulates the spread of measles in a highschool. The highschool has 1000 students, and the simulation runs for 60 days with one index case. 
We can check that transitions between states looking at the transition matrix:

```
Transition Probabilities:
 - Susceptible              1.00  0.00  0.00  0.00  0.00  0.00  0.00  0.00  0.00
 - Exposed                  0.00  0.68  0.19  0.00  0.00  0.11  0.00  0.02  0.00
 - Prodromal                0.00  0.00  0.70  0.26  0.00  0.00  0.00  0.04  0.00
 - Rash                     0.00  0.00  0.00  0.48  0.52  0.00  0.00  0.00  0.00
 - Isolated                 0.00  0.00  0.00  0.00  0.72  0.00  0.00  0.00  0.28
 - Quarantined Exposed      0.00  0.00  0.00  0.00  0.00  0.86  0.00  0.14  0.00
 - Quarantined Susceptible  0.08  0.00  0.00  0.00  0.00  0.00  0.92  0.00  0.00
 - Quarantined Infectious   0.00  0.00  0.05  0.00  0.22  0.00  0.00  0.73  0.00
 - Recovered  
```

From it, we see:

- Exposed can move to prodromal, quarantined exposed, or quarantined infectious.
- Prodromal moves to Rash or Quarantined Prodromal (which is infectious).
- Rash moves to Isolated (and triggers contact tracing).
- Isolated moves to Recovered (full immunity).
- Quarantined Exposed either moves to quarantined infectious or to isolated. It can moved to exposed, but since the quarantine period is 14 days, we don't see any case.
- Quarantined Infectious moves to prodromal (if exits quarantine) or to isolated (if rash detected), in which case contact tracing is triggered.


```{r}
#| label: reading-history
library(data.table)
histories <- lapply(list.files(
    "highschool-out", pattern="total_hist", full.names = TRUE),
   \(fn) {
      cbind(fn = fn, data.table::fread(fn))
   }) |> rbindlist()


# Figuring out the outbreak size
histories[state != "Susceptible" & date == max(date)]
```